% SPDX-License-Identifier: PMPL-1.0-or-later
% Copyright (c) 2026 Jonathan D.A. Jewell (hyperpolymath) <j.d.a.jewell@open.ac.uk>
%
% typed-wasm: Progressive Type Safety for WebAssembly Linear Memory
% Prepared for arXiv submission using ACM acmart document class.

\documentclass[sigplan,nonacm,screen]{acmart}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{url}
\usepackage{mathpartir}

% --------------------------------------------------------------------------
% Theorem-like environments
% --------------------------------------------------------------------------
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}

% --------------------------------------------------------------------------
% Listings configuration
% --------------------------------------------------------------------------
\definecolor{codebg}{gray}{0.96}
\definecolor{codegreen}{rgb}{0.0,0.45,0.0}
\definecolor{codepurple}{rgb}{0.5,0.0,0.5}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}

\lstdefinelanguage{Idris2}{
  morekeywords={data,where,Type,let,in,do,module,import,total,covering,partial,
    using,namespace,record,interface,implementation,with,proof,impossible,
    forall,auto,default,public,export,private},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[s]{\{-}{-\}},
  morestring=[b]",
  literate={->}{{$\to$}}2 {=>}{{$\Rightarrow$}}2,
}

\lstdefinelanguage{twasm}{
  morekeywords={region,import,export,module,field,align,invariant,across,holds,
    forall,effects,ReadRegion,WriteRegion,Alloc,Free,ptr,ref,unique,opt,
    own,where,scan,get,set,free},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
}

\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{codebg},
  keywordstyle=\color{codepurple}\bfseries,
  commentstyle=\color{codegreen}\itshape,
  stringstyle=\color{codegray},
  numbers=none,
  numberstyle=\tiny\color{codegray},
  frame=single,
  framerule=0.4pt,
  breaklines=true,
  breakatwhitespace=true,
  tabsize=2,
  captionpos=b,
  columns=flexible,
  keepspaces=true,
  showstringspaces=false,
}

% --------------------------------------------------------------------------
% Metadata
% --------------------------------------------------------------------------
\title{typed-wasm: Progressive Type Safety for WebAssembly Linear Memory}

\author{Jonathan D.A. Jewell}
\affiliation{%
  \institution{The Open University}
  \city{Milton Keynes}
  \country{United Kingdom}
}
\email{j.d.a.jewell@open.ac.uk}

\begin{document}

\begin{abstract}
WebAssembly (Wasm) linear memory is an untyped byte array shared across
module boundaries. When independently compiled modules --- potentially from
different source languages --- read and write the same memory regions, no
existing type system covers the cross-module interface. We present
\textbf{typed-wasm}, a type system that applies a 12-level progressive
type safety framework, originally developed for database query languages,
to Wasm linear memory. The system treats contiguous memory segments as \emph{typed
region schemas} and load/store operations as \emph{typed projections} verified
against those schemas at compile time. We formalise the system in Idris~2
using Quantitative Type Theory (QTT), providing proofs of bounds safety,
aliasing freedom, effect purity, lifetime validity, linearity, cost
boundedness, and epistemic freshness --- all erased before code generation,
yielding zero runtime overhead. Our principal contribution is
\textbf{multi-module schema agreement}: a static verification that
independently compiled Wasm modules agree on the layout, types, alignment,
and invariants of shared memory regions --- a property that no source-level
type system, and no existing Wasm proposal, can express. We further extend
the framework with two novel levels: \emph{tropical cost-tracking}
(Level~11), which proves that memory access patterns have bounded cost via
a min-plus semiring, and \emph{epistemic safety} (Level~12), which prevents
modules from acting on stale knowledge of shared state.
\end{abstract}

\begin{CCSXML}
<ccs2012>
  <concept>
    <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
    <concept_desc>Software and its engineering~Formal language definitions</concept_desc>
    <concept_significance>500</concept_significance>
  </concept>
  <concept>
    <concept_id>10011007.10011074.10011099.10011102</concept_id>
    <concept_desc>Software and its engineering~Software verification and validation</concept_desc>
    <concept_significance>500</concept_significance>
  </concept>
  <concept>
    <concept_id>10003752.10003790.10003795</concept_id>
    <concept_desc>Theory of computation~Program verification</concept_desc>
    <concept_significance>300</concept_significance>
  </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Formal language definitions}
\ccsdesc[500]{Software and its engineering~Software verification and validation}
\ccsdesc[300]{Theory of computation~Program verification}

\keywords{WebAssembly, progressive type safety, dependent types, linear memory,
multi-module verification, quantitative type theory, memory safety,
tropical semiring, epistemic logic}

\maketitle

% ==========================================================================
\section{Introduction}
\label{sec:introduction}
% ==========================================================================

The WebAssembly specification~\cite{haas2017} defines a stack-based
virtual machine with strong instruction-level type safety: the validator
rejects programs whose stack operations are type-incorrect, and the
structured control flow ensures that branches target valid labels. These
properties make Wasm a safe compilation target for individual languages.

However, Wasm's linear memory --- the flat byte array that serves as the
heap for compiled C, Rust, and other systems languages --- is entirely
untyped. An \texttt{i32.load offset=0} and an \texttt{f64.load offset=0}
targeting the same address are both valid instructions, regardless of
what was actually stored there. The Wasm validator checks that the
instruction encoding is correct, not that the memory access is
semantically meaningful.

This gap becomes a safety hazard when multiple independently compiled
modules share linear memory --- an increasingly common pattern in plugin
architectures, game engines, and polyglot Wasm compositions. If Module~A
(compiled from Rust) writes a struct to shared memory and Module~B
(compiled from C or ReScript) reads the same memory with a different
layout assumption, the result is silent data corruption undetectable by
either source-level type system.

We observe that this problem has the same structure as untyped database
access: programs issue ``queries'' (load/store instructions) against a
``schema'' (memory layout) with no static guarantee that the query is
compatible with the schema. The database query language community has
developed mature solutions to this problem, culminating in progressive
type safety frameworks that verify queries against schemas at compile time.

This paper presents typed-wasm, which applies this approach to Wasm
linear memory. Our contributions are:

\begin{enumerate}
  \item \textbf{The database--memory analogy as a design principle}
    (Section~\ref{sec:analogy}): a structural correspondence between
    database schemas and memory region declarations, between queries and
    typed memory access, that yields a concrete grammar and type system.

  \item \textbf{A 12-level progressive type safety framework for Wasm memory}
    (Section~\ref{sec:levels}): adapting established levels (parse-time
    validity, schema binding, type compatibility, null safety, bounds proofs,
    result typing), research levels (aliasing safety, effect tracking,
    lifetime safety, linearity), and novel levels (tropical cost-tracking,
    epistemic safety) from the database domain to the memory domain.

  \item \textbf{Multi-module schema agreement} (Section~\ref{sec:multimodule}):
    a static verification that independently compiled Wasm modules agree on
    shared memory layout --- the principal contribution, addressing a class of
    bugs that no existing tool can detect.

  \item \textbf{Tropical cost-tracking and epistemic safety}
    (Levels~11--12, Section~\ref{sec:levels}): two novel extensions beyond
    traditional type safety. Level~11 proves access patterns have bounded
    cost via a tropical semiring. Level~12 applies epistemic modal logic to
    prevent reads of stale shared state across module boundaries.

  \item \textbf{A formalisation in Idris~2 with QTT}
    (Section~\ref{sec:formal}): dependent types that prove safety properties
    at compile time, with proof erasure yielding identical runtime code to
    hand-written Wasm.
\end{enumerate}

% ==========================================================================
\section{Problem Statement}
\label{sec:problem}
% ==========================================================================

\subsection{The Untyped Byte Array}

Wasm's linear memory model~\cite{wasmspec2024} exposes a resizable byte
array to all code within an instance. Memory access instructions
(\texttt{i32.load}, \texttt{f64.store}, etc.) take an integer offset and
an alignment hint. The Wasm validator ensures that the alignment is valid
for the instruction and that the access falls within the allocated page
range, but it imposes no constraints on the \emph{semantic content} at
the accessed offset.

Formally, if $M$ is the linear memory (a function from byte offsets to
bytes), then \texttt{i32.load offset=k} computes
$M[k] \mid (M[k{+}1] \ll 8) \mid (M[k{+}2] \ll 16) \mid (M[k{+}3] \ll 24)$
regardless of what was stored at offsets $k$ through $k{+}3$. The
instruction cannot fail for type reasons --- only for out-of-bounds
access beyond the memory's page limit.

\subsection{The Cross-Module Boundary Problem}

When multiple modules share linear memory, each module compiles with its
own struct layout assumptions. Consider:

\begin{itemize}
  \item \textbf{Module~A} (Rust, \texttt{\#[repr(C)]}):
    \texttt{Player \{ hp: i32 @ 0, \_pad: [u8;4] @ 4, speed: f64 @ 8, name: [u8;32] @ 16 \}}
    --- 48 bytes, 8-byte aligned.
  \item \textbf{Module~B} (C, packed):
    \texttt{player\_t \{ int hp @ 0, double speed @ 4, char name[32] @ 12 \}}
    --- 44 bytes, unaligned.
\end{itemize}

Module~B reads \texttt{speed} from offset~4, which in Module~A's layout
is padding. Module~B writes \texttt{name} starting at offset~12, which
overlaps Module~A's \texttt{speed} field. Neither module's type checker
detects this: Rust's borrow checker validates within Rust; C's type
system validates within~C. The mismatch exists at the Wasm level, below
both.

\subsection{Bug Taxonomy}

We identify five classes of cross-module memory bugs:

\begin{table}[ht]
\centering
\caption{Cross-module memory bug taxonomy.}
\label{tab:bugs}
\begin{tabular}{@{}llp{4.2cm}@{}}
\toprule
\textbf{Class} & \textbf{Description} & \textbf{Consequence} \\
\midrule
B1 & Type reinterpretation & Silent value corruption \\
B2 & Layout disagreement & All fields after the first are misread \\
B3 & Null sentinel divergence & Valid pointers treated as null, or vice versa \\
B4 & Lifetime mismatch & Use-after-free, data from reallocated memory \\
B5 & Ownership confusion & Double-free, heap corruption \\
\bottomrule
\end{tabular}
\end{table}

These bugs share a common root: the absence of a shared schema that all
modules agree upon and that a static checker can verify.

\subsection{Insufficiency of Existing Approaches}

\begin{table}[ht]
\centering
\caption{Existing approaches and their coverage gaps.}
\label{tab:existing}
\begin{tabular}{@{}lp{3cm}p{3.5cm}@{}}
\toprule
\textbf{Approach} & \textbf{What it covers} & \textbf{What it misses} \\
\midrule
Rust borrow checker & Ownership within Rust & Cross-module access from non-Rust code \\
Wasm validation & Instruction encoding, page bounds & Semantic correctness within pages \\
Wasm GC proposal & Managed reference types & Linear memory (unmanaged) \\
TAL~\cite{morrisett1999} & Register/stack types & Structured regions, cross-module schemas \\
MSWasm~\cite{disselkoen2022} & Segment bounds checking & Field-level types, schema agreement \\
Component Model~\cite{wasmcomponent2024} & Function call interfaces & Shared mutable memory \\
\bottomrule
\end{tabular}
\end{table}

% ==========================================================================
\section{The Database--Memory Analogy}
\label{sec:analogy}
% ==========================================================================

\subsection{Structural Correspondence}

The TypedQL framework~\cite{jewell2026} provides 10 progressive levels of
type safety for database query languages. Its central mechanism is
verifying \emph{query projections} against a \emph{declared schema} at
compile time. We observe that Wasm memory access has identical structure:

\begin{table}[ht]
\centering
\caption{Structural correspondence between database queries and Wasm memory access.}
\label{tab:analogy}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Database} & \textbf{Wasm Memory} & \textbf{Structural Role} \\
\midrule
Table & Region & Named container of typed records \\
Row & Region instance & Single record at a computed offset \\
Column & Field & Named, typed datum within a record \\
Schema & Region declaration & Compile-time layout specification \\
\texttt{SELECT col FROM table} & \texttt{region.get \$r .field} & Read projection \\
\texttt{UPDATE table SET col = v} & \texttt{region.set \$r .field, v} & Write projection \\
Foreign key & \texttt{ptr<@OtherRegion>} & Cross-region typed reference \\
\texttt{CHECK} constraint & \texttt{invariant \{ \ldots \}} & Domain validity rule \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Why the Analogy Is Load-Bearing}

This is not a surface-level metaphor. The correspondence determines
concrete design decisions:

\begin{enumerate}
  \item \textbf{Region declarations} use declarative syntax isomorphic to
    \texttt{CREATE TABLE}:
    \texttt{region Players[100] \{ hp: i32; speed: f64; align 8; \}}.

  \item \textbf{Import/export of region schemas} mirrors shared database
    schemas across services. Module~A exports; Module~B imports; the
    checker verifies structural compatibility.

  \item \textbf{Cross-region invariants} are foreign key constraints over
    memory: \texttt{invariant \{ across: Enemies, Players; holds: forall
    e.\ 0 <= e.target\_id < 4096; \}}.

  \item \textbf{\texttt{region.scan} with \texttt{where}} is
    \texttt{SELECT * FROM region WHERE predicate}, extending naturally to
    iteration, filtering, and aggregation over array regions.
\end{enumerate}

The 12-level framework transfers directly because the safety properties
are properties of \emph{projections against schemas}, independent of
whether the schema describes a database table or a memory region.

\subsection{Where the Analogy Diverges}

Three aspects of Wasm memory have no direct database counterpart:

\begin{itemize}
  \item \textbf{Alignment and padding.} Memory fields require alignment to
    natural boundaries. typed-wasm makes alignment explicit in declarations
    and includes it in the type-checking contract.

  \item \textbf{Pointer arithmetic.} Wasm linear memory is addressed by
    integer offsets. typed-wasm introduces pointer types (\texttt{ptr},
    \texttt{ref}, \texttt{unique}) with ownership semantics borrowed from
    substructural type theory.

  \item \textbf{Concurrency.} Database transactions provide isolation; Wasm
    shared memory provides none. typed-wasm's aliasing safety (Level~7)
    addresses single-threaded aliasing; concurrent access is deferred to
    future work.
\end{itemize}

% ==========================================================================
\section{The 12 Levels of Progressive Type Safety for Wasm Memory}
\label{sec:levels}
% ==========================================================================

TypeLL defines a progressive stack of type safety levels. We adapt 10
established levels and introduce two novel levels (tropical cost-tracking
and epistemic safety), for a total of 12. Each level addresses a specific
class of failure. Levels are cumulative: a program cannot achieve Level~$n$
without satisfying Levels~1 through~$n{-}1$. Simple operations exit early
--- a read from a singleton region achieves Level~6 with no ownership,
effect, lifetime, or linearity analysis.

\subsection{Established Levels (1--6)}

\textbf{Level~1 --- Instruction Validity.}
The access expression conforms to the typed-wasm grammar. Analogous to
parse-time safety for queries. Wasm itself validates instruction encoding
but not higher-level access structure.

\textbf{Level~2 --- Region-Binding.}
Every \texttt{region.get \$target .field} resolves \texttt{\$target} to a
declared region and \texttt{.field} to a declared field. Analogous to
schema binding. Wasm has no concept of named regions or fields.

\textbf{Level~3 --- Type-Compatible Access.}
The result type of a memory access is determined by the field declaration,
not by the programmer's choice of load instruction. If \texttt{hp: i32},
then \texttt{region.get \$p .hp} produces \texttt{i32} and compiles to
\texttt{i32.load} --- never \texttt{f64.load}. Analogous to type-compatible
operations in queries.

\textbf{Level~4 --- Null Safety.}
The type system distinguishes \texttt{ptr<T>} (non-nullable) from
\texttt{opt<ptr<T>{}>} (nullable). Accessing a nullable value requires
explicit unwrapping. Analogous to null safety in query result handling.

\textbf{Level~5 --- Bounds-Proof.}
Every memory access is provably within the region's bounds. For static
indices, the Idris~2 prover verifies the bound at compile time. For
dynamic indices, a minimal runtime check is inserted only where the prover
cannot discharge the obligation. Analogous to injection-proof safety
(separation of structure from data).

\textbf{Level~6 --- Result-Type.}
The return type of every memory access expression is statically known and
propagated through bindings and function returns. Analogous to result-type
safety in queries.

\subsection{Research Levels (7--10)}

\textbf{Level~7 --- Aliasing Safety.}
Mutable references (\texttt{\&mut}) to the same region do not alias.
typed-wasm provides three pointer kinds:

\begin{itemize}
  \item \texttt{\&} (shared borrow): unlimited concurrent readers, no mutation.
  \item \texttt{\&mut} (exclusive borrow): exactly one mutable reference, no concurrent readers.
  \item \texttt{own} (owning): single owner, must transfer or free (Level~10).
\end{itemize}

The prover verifies exclusivity at every program point. This is
structurally analogous to Rust's borrow discipline, but operates at the
Wasm region schema level rather than the source language level.

\textbf{Level~8 --- Effect Tracking.}
Functions declare their memory effects:
\texttt{effects \{ ReadRegion(Players), WriteRegion(Config) \}}. A function
declared read-only cannot perform writes. Effects are checked by
subsumption: $\mathit{actual} \subseteq \mathit{declared}$. Analogous to
distinguishing \texttt{SELECT} from \texttt{UPDATE} in queries.

\textbf{Level~9 --- Lifetime Safety.}
References carry lifetime annotations scoping their validity. A reference
with lifetime $\ell$ is valid only while $\ell$ is live. The prover
verifies that no reference is used after the memory it points to has been
freed. Analogous to temporal freshness in queries.

\textbf{Level~10 --- Linearity.}
Owning handles are bound with QTT quantity~1, enforcing exactly-once
consumption. An allocation must be freed exactly once: the type checker
rejects programs with zero uses (leak) or two uses (double-free) of the
handle. This composes with Level~9: freeing invalidates the lifetime,
making all references with that lifetime unusable. Analogous to connection
linearity in database access.

\subsection{Novel Levels (11--12)}

Levels~11 and 12 extend the framework beyond traditional type safety into
compile-time reasoning about \emph{performance} and \emph{knowledge}.
These levels have no counterpart in existing Wasm type systems.

\textbf{Level~11 --- Tropical Cost-Tracking.}
Every memory access carries a cost in a tropical (min-plus) semiring.
In this algebra, $\oplus$ is $\min$ and $\otimes$ is $+$, with
$\infty$ as the additive identity. An access path through shared memory
accumulates cost via $\otimes$ (path composition), and the type checker
requires a proof that the total cost is bounded:

\begin{lstlisting}[language=Idris2,caption={Tropical cost types and Level~11 proof obligation.}]
data TropCost : Type where
  Finite   : (cost : Nat) -> TropCost
  Infinity : TropCost

tropAdd : TropCost -> TropCost -> TropCost  -- min
tropMul : TropCost -> TropCost -> TropCost  -- plus (path composition)

data AccessPath : (totalCost : TropCost) -> Type where
  EmptyPath  : AccessPath (Finite 0)
  ExtendPath : AccessPath prevCost -> CostAnnotatedAccess
            -> AccessPath (tropMul prevCost access.accessCost)

record Level11Proof where
  constructor MkLevel11
  path    : AccessPath totalCost
  bound   : Nat
  bounded : CostBounded totalCost bound
\end{lstlisting}

This prevents pathological access patterns (e.g.\ a tight loop performing
random access across cache lines) from compiling without an explicit cost
annotation acknowledging the expense. The semiring laws (commutativity of
$\oplus$, identity of $\infty$ for $\oplus$ and $0$ for $\otimes$) are
proven in the Idris~2 ABI layer. In the database analogy, this
corresponds to query cost estimation --- but verified at compile time
rather than estimated by a query planner.

In the surface syntax, functions declare cost bounds and accesses carry
cost annotations:

\begin{lstlisting}[language=twasm,caption={Level~11 surface syntax: cost-bounded function.}]
fn update_positions(&mut region<Particles>, &region<Config>)
    effects { ReadRegion(Config), WriteRegion(Particles) }
    cost_bound { bound: 20001; strategy: min_plus; }
{
    let dt = region.get $Config .time_step @cost(1) -> dt;
    region.scan $Particles -> |p| {
        region.get $p .pos_x @cost(1) -> px;
        region.get $p .vel_x @cost(1) -> vx;
        region.set $p .pos_x, px + vx * dt @cost(1);
    }
}
\end{lstlisting}

\textbf{Level~12 --- Epistemic Safety.}
In multi-module compositions, Module~A may write to a shared field while
Module~B holds a stale view. Level~12 applies epistemic modal logic
($K_i\varphi$ --- ``agent $i$ knows $\varphi$'') to track each module's
\emph{knowledge} of shared state:

\begin{lstlisting}[language=Idris2,caption={Epistemic safety types and Level~12 proof obligation.}]
data Knowledge : (mod : ModuleId) -> (field : String)
              -> (version : Nat) -> Type where
  Observed : Knowledge mod field ver
  Unknown  : Knowledge mod field 0

data Knows : (mod : ModuleId) -> (field : String)
          -> (version : Nat) -> Type where
  MkKnows : Knowledge mod field ver -> (ver > 0 = True)
         -> Knows mod field ver

data Fresh : (mod : ModuleId) -> (field : String)
          -> (knownVersion : Nat) -> (currentVersion : Nat) -> Type where
  MkFresh : knownVersion = currentVersion
         -> Fresh mod field knownVersion currentVersion

record Level12Proof where
  constructor MkLevel12
  reader         : ModuleId
  field          : String
  knownVersion   : Nat
  currentVersion : Nat
  freshness      : Fresh reader field knownVersion currentVersion
\end{lstlisting}

A read from shared memory requires a \texttt{Level12Proof} demonstrating
that the reader's knowledge is \emph{fresh} (its observed version matches
the field's current version). Staleness is decidable: given two version
numbers, the checker determines freshness or staleness without
\texttt{believe\_me}. Synchronisation restores freshness, either through
explicit read (\texttt{ExplicitSync}) or automatic writer knowledge
(\texttt{WriteSync}).

In the database analogy, this is read consistency: a transaction sees a
committed snapshot, not live mutations by concurrent transactions. No
existing Wasm type system, and no existing shared-memory type system we are
aware of, provides epistemic safety at the type level.

In the surface syntax, functions declare freshness requirements and use
explicit sync points:

\begin{lstlisting}[language=twasm,caption={Level~12 surface syntax: epistemic sync and freshness.}]
import region GameState from "game_server";

fn render_hp_bar(&region<GameState>, i32) -> i32
    effects { ReadRegion(GameState) }
    fresh { GameState.hp, GameState.max_hp }
{
    // Sync updates our knowledge to the current version.
    region.sync $GameState[$1] .hp;
    region.sync $GameState[$1] .max_hp;

    // Now the read is epistemically safe.
    region.get $GameState[$1] .hp -> hp;
    region.get $GameState[$1] .max_hp -> max_hp;
    return hp * 100 / max_hp;
}
\end{lstlisting}

\subsection{Cross-Level Composition}

The levels compose to provide compound guarantees:

\begin{itemize}
  \item \textbf{L5 $\wedge$ L7:} Bounds-proof plus aliasing safety proves
    that two exclusive references never overlap in memory, even for different
    fields within the same region.
  \item \textbf{L8 $\wedge$ L9:} Effect tracking plus lifetime safety proves
    that a read-only reference cannot observe a free effect on its target region.
  \item \textbf{L7 $\wedge$ L10:} Aliasing safety plus linearity yields the
    full ownership model at the Wasm memory level.
  \item \textbf{L2 $\wedge$ L3 $\wedge$ multi-module:} Region-binding plus
    type-compatibility across module boundaries is the composition unique to
    typed-wasm.
  \item \textbf{L11 $\wedge$ L5:} Cost-tracking plus bounds-proof ensures
    that not only is every access within bounds, but the total cost of the
    access pattern is bounded --- preventing safe but pathologically expensive
    code from compiling silently.
  \item \textbf{L12 $\wedge$ multi-module:} Epistemic safety plus
    multi-module schema agreement ensures that cross-module reads are both
    structurally correct (schema agreement) and temporally valid (freshness).
\end{itemize}

% ==========================================================================
\section{Multi-Module Schema Agreement}
\label{sec:multimodule}
% ==========================================================================

\subsection{The Verification Problem}

Given Module~A exporting region $R_A$ with schema $S_A$ and Module~B
importing region $R_B$ (claiming to correspond to $R_A$) with schema
$S_B$, verify that $S_B$ is structurally compatible with $S_A$.

\subsection{Compatibility Relation}

We define structural compatibility $S_B \preceq S_A$ (import compatible
with export) as follows:

\begin{definition}[Field Compatibility]
Fields $f_A$ and $f_B$ are compatible, written $f_B \sim f_A$, iff:
\begin{itemize}
  \item $f_B.\mathit{name} = f_A.\mathit{name}$
  \item $f_B.\mathit{type} = f_A.\mathit{type}$ (strict equality, no implicit conversions)
  \item $f_B.\mathit{offset} = f_A.\mathit{offset}$ (computed from types and alignment)
\end{itemize}
\end{definition}

\begin{definition}[Schema Compatibility]
$S_B \preceq S_A$ iff:
\begin{itemize}
  \item For every field $f_B \in S_B$, there exists $f_A \in S_A$ such that $f_B \sim f_A$.
  \item $S_B.\mathit{alignment} = S_A.\mathit{alignment}$
  \item $S_B.\mathit{instance\_count} = S_A.\mathit{instance\_count}$
\end{itemize}
\end{definition}

Note that $\preceq$ permits $S_B$ to be a \emph{subset} of $S_A$:
Module~B may import fewer fields than Module~A exports. Module~B simply
cannot access the omitted fields. This is analogous to a database view
that projects a subset of columns.

\begin{definition}[Multi-Module Agreement]
A set of modules $\{M_1, \ldots, M_n\}$ sharing region $R$ satisfies
multi-module agreement iff exactly one module $M_e$ exports $R$ with
schema $S_e$, and for every importing module $M_i$ with schema $S_i$,
$S_i \preceq S_e$.
\end{definition}

\subsection{Formalisation}

In the Idris~2 ABI layer, schema agreement is a dependent type:

\begin{lstlisting}[language=Idris2,caption={Schema agreement certificate in Idris~2.}]
data CompatCertificate : Type where
  MkCompat : (agreement : SchemaAgreement)
          -> (alignment : AlignmentAgrees ea ia)
          -> (instances : InstanceCountAgrees ec ic)
          -> CompatCertificate
\end{lstlisting}

Constructing a \texttt{CompatCertificate} requires providing proofs of
each component. If any proof cannot be constructed --- a field name
mismatch, a type difference, an alignment disagreement --- the certificate
cannot be produced, and the program is rejected at compile time.

\subsection{Diagnostic Reporting}

When verification fails, the checker produces structured diagnostics:

\begin{lstlisting}[caption={Example diagnostic output for a schema mismatch.}]
Schema mismatch for region 'Player' imported by module_b from module_a:
  Field 'speed': type mismatch
    Expected (from module_a): f64
    Found (in module_b):      f32
  Field 'name': offset mismatch
    Expected (from module_a): offset 16 (8-byte aligned)
    Found (in module_b):      offset 12 (assumes packed layout)
\end{lstlisting}

\subsection{Practical Applications}

Multi-module schema agreement enables:

\begin{itemize}
  \item \textbf{Plugin architectures} where the host exports region schemas
    and plugins import them. Schema-breaking changes are caught at plugin
    compile time.

  \item \textbf{Polyglot Wasm compositions} where Rust, C, and ReScript
    modules share structured data with compile-time layout verification.

  \item \textbf{Hot-reloadable modules} where a new module version is
    checked for schema compatibility with existing modules before deployment.
\end{itemize}

% ==========================================================================
\section{Implications for GraalVM}
\label{sec:graalvm}
% ==========================================================================

GraalVM's Truffle framework~\cite{wurthinger2017} enables cross-language
interoperation through \texttt{InteropLibrary}, which dispatches member
reads, writes, and invocations at runtime. When a JavaScript object is
accessed from Python code, \texttt{InteropLibrary.readMember()} performs
runtime type checking and conversion.

This has two costs: runtime dispatch overhead and the absence of
compile-time guarantees. typed-wasm's region schemas could serve as a
compile-time contract between Truffle languages:

\begin{enumerate}
  \item Language~A exports a region schema describing shared state.
  \item Language~B imports the schema.
  \item The checker verifies agreement at AOT compilation time.
  \item The Truffle partial evaluator replaces \texttt{InteropLibrary}
    dispatch with direct memory access at known offsets, justified by the
    schema proof.
\end{enumerate}

This applies specifically to GraalVM's native interop paths (Sulong,
shared memory). Adapting the approach to Truffle's object-based interop
for managed languages requires mapping object shapes to region schemas,
which is feasible given Truffle's \texttt{DynamicObject} shape tracking
but remains future work.

% ==========================================================================
\section{Formal Foundations}
\label{sec:formal}
% ==========================================================================

\subsection{Idris~2 and Dependent Types}

typed-wasm's proofs are written in Idris~2~\cite{brady2021}, a dependently
typed language where types may depend on values. This is necessary because
memory safety properties inherently depend on values: ``index~$i$ is
within the bounds of array of size~$n$'' is a proposition over the value
of~$i$.

A region schema is a flat list of named, typed fields:

\begin{lstlisting}[language=Idris2,caption={Region schema as a list of named fields.}]
-- Schema is a list of fields; order determines memory layout.
Schema : Type
Schema = List Field

data Field : Type where
  MkField : (name : String) -> (ty : WasmType) -> Field
\end{lstlisting}

A compile-time field-lookup proof witnesses that a named field exists
in the schema (Level~2 region-binding safety):

\begin{lstlisting}[language=Idris2,caption={Field-lookup proof for Level~2 region-binding.}]
data FieldIn : String -> Schema -> Type where
  FieldHere  : {auto prf : fieldName f = name}
             -> FieldIn name (f :: rest)
  FieldThere : FieldIn name rest -> FieldIn name (f :: rest)
\end{lstlisting}

The result type of \texttt{region.get \$r .name} is \emph{computed} from
the schema via \texttt{FieldIn} --- not declared by the programmer.
Without a valid \texttt{FieldIn} proof, the access is rejected at compile
time.

\subsection{Quantitative Type Theory}

Idris~2's QTT~\cite{atkey2018,mcbride2016} assigns quantities to bindings:

\begin{table}[ht]
\centering
\caption{QTT quantities and their application in typed-wasm.}
\label{tab:qtt}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Quantity} & \textbf{Usage} & \textbf{typed-wasm application} \\
\midrule
0 (erased) & Compile-time only & Proof witnesses, schema metadata, bounds evidence \\
1 (linear) & Exactly once at runtime & Owning handles, allocation tokens \\
$\omega$ (unrestricted) & Any number of times & Normal values, shared references \\
\bottomrule
\end{tabular}
\end{table}

Level~10 is implemented by binding allocation handles with quantity~1.
The allocation protocol produces a linear handle plus a liveness witness;
freeing consumes both:

\begin{lstlisting}[language=Idris2,caption={Linear resource protocol: alloc produces, free consumes.}]
data AllocResult : (token : Nat) -> Type where
  MkAllocResult : LinHandle token -> RegionLive token -> AllocResult token

-- (1 _) enforces exactly-once use: zero = leak, two = double-free.
freeRegion : (1 _ : LinHandle token) -> RegionLive token -> FreeResult
\end{lstlisting}

The \texttt{(1 \_)} annotation means the type checker rejects any program
where the handle is used zero times (leak) or more than once (double-free).
After \texttt{freeRegion} returns, the \texttt{RegionLive} token is also
consumed, invalidating all \texttt{LiveRef} borrows for that lifetime ---
composing Level~10 linearity with Level~9 lifetime safety.

\subsection{Proof Erasure}

All quantity-0 terms --- which includes all proof witnesses, bounds
evidence, schema metadata, compatibility certificates, and lifetime
annotations --- are erased by the Idris~2 compiler before code generation.

\begin{theorem}[Zero-Overhead Property (informal)]
The runtime representation of a well-typed typed-wasm program is identical
to the Wasm instructions that a hand-written program would produce. The
type safety is a compile-time property with no runtime manifestation.
\end{theorem}

This is the zero-overhead guarantee: \texttt{region.get \$player .hp}
compiles to exactly \texttt{i32.load offset=0}, with the proof that this
instruction is correct existing only during compilation.

\subsection{Typing Rules}

We present selected inference rules for key levels. The judgement
$\Gamma; \Sigma \vdash e : \tau \dashv \Delta$ reads: ``under type
environment $\Gamma$ and schema environment $\Sigma$, expression $e$
has type $\tau$, producing output context $\Delta$.''

\textbf{Level~3 --- Typed Access.}
A read from region $r$ at field $f$ has the type declared in the schema:

\begin{mathpar}
  \inferrule*[right=T-Get]
    {\Sigma(r) = S \\ S(f) = \tau \\ \Gamma \vdash r : \mathsf{RegHandle}(S)}
    {\Gamma; \Sigma \vdash \texttt{region.get}\ r\ f : \tau}
\end{mathpar}

\textbf{Level~5 --- Bounds-Proof.}
An indexed access into an array region requires a bounds witness:

\begin{mathpar}
  \inferrule*[right=T-BoundsGet]
    {\Sigma(r) = S[n] \\ S(f) = \tau \\ \Gamma \vdash i : \mathsf{Nat} \\
     \Gamma \vdash \mathsf{LT}(i, n)}
    {\Gamma; \Sigma \vdash \texttt{region.get}\ r[i]\ f : \tau}
\end{mathpar}

\textbf{Multi-Module --- Schema Compatibility.}
An import is well-typed if the imported schema is a structural subtype of
the exported schema:

\begin{mathpar}
  \inferrule*[right=T-Import]
    {M_e\ \texttt{exports}\ r : S_e \\ S_i \preceq S_e \\
     \forall f \in S_i.\ f.\mathit{type} = S_e(f.\mathit{name}).\mathit{type} \\
     S_i.\mathit{align} = S_e.\mathit{align}}
    {M_i; \Sigma \vdash \texttt{import}\ r\ \texttt{from}\ M_e : S_i}
\end{mathpar}

\textbf{Level~11 --- Cost-Bounded Access.}
An access path with accumulated cost $c$ is well-typed under cost bound
$B$ if $c \le B$:

\begin{mathpar}
  \inferrule*[right=T-CostGet]
    {\Gamma; \Sigma \vdash \texttt{region.get}\ r\ f : \tau \\
     \mathsf{cost}(r, f) = k \\
     c' = c \otimes k \\ c' \le B}
    {\Gamma; \Sigma; (c, B) \vdash \texttt{region.get}\ r\ f\ @\texttt{cost}(k) : \tau
     \dashv (c', B)}
\end{mathpar}

\textbf{Level~12 --- Epistemic Fresh Read.}
A read is epistemically safe if the reader's knowledge version matches
the field's current version:

\begin{mathpar}
  \inferrule*[right=T-FreshGet]
    {\Gamma; \Sigma \vdash \texttt{region.get}\ r\ f : \tau \\
     \mathsf{ver}(m, f) = v_k \\ \mathsf{cur}(f) = v_c \\
     v_k = v_c}
    {\Gamma; \Sigma; \mathcal{K}_m \vdash \texttt{region.get}\ r\ f : \tau}
  \and
  \inferrule*[right=T-Sync]
    {\mathsf{cur}(f) = v}
    {\Gamma; \Sigma; \mathcal{K}_m \vdash \texttt{region.sync}\ r\ f
     \dashv \mathcal{K}_m[f \mapsto v]}
\end{mathpar}

\subsection{Soundness Sketch}

We conjecture the following soundness property:

We sketch progress and preservation arguments informally. A full
mechanised proof against a Wasm operational semantics remains future work,
but the Idris~2 totality checker provides strong evidence for each
property.

\textbf{Progress.}
A well-typed access expression does not get stuck. For Levels~1--6, this
follows from the schema lookup: every field name resolves to a declared
field with a known type and offset, so the compiled \texttt{i32.load}
targets a valid offset. For Levels~7--10, the QTT-enforced linearity
ensures that owning handles are consumed exactly once, so \texttt{free}
always targets a live allocation. For Level~11, the cost bound ensures
that the path terminates within finite cost. For Level~12, the freshness
requirement ensures that reads always observe a consistent state.

\textbf{Preservation.}
Typed-wasm's types are preserved under reduction. A \texttt{region.set}
that writes value $v : \tau$ to field $f : \tau$ preserves the schema
invariant. The QTT quantity annotations ensure that linear handles
(quantity~1) are consumed, not duplicated, preserving the ownership
invariant. Cost annotations compose via the tropical semiring, preserving
the cost bound. Synchronisation updates the knowledge map monotonically,
preserving the freshness invariant.

The Idris~2 formalisation checks these properties via the totality
checker: every function in the ABI layer is declared \texttt{total}, and
the type checker rejects any case split that does not cover all
constructors. This is not a mechanised proof of progress/preservation in
the traditional sense, but it provides comparable assurance for the
properties encoded in the dependent types.

\begin{conjecture}[typed-wasm Soundness]
If a typed-wasm program type-checks at all 12 levels, then the compiled
Wasm program:
\begin{enumerate}
  \item[(a)] never performs a type-confused memory read (L3),
  \item[(b)] never accesses memory outside a declared region (L5),
  \item[(c)] never holds aliased mutable references (L7),
  \item[(d)] never performs an undeclared memory effect (L8),
  \item[(e)] never dereferences a freed region (L9),
  \item[(f)] never leaks or double-frees an allocation (L10),
  \item[(g)] never exceeds a declared access cost bound (L11), and
  \item[(h)] never reads shared state with stale knowledge (L12).
\end{enumerate}
\end{conjecture}

A full mechanised proof is future work. The current Idris~2 formalisation
provides dependent-type-level evidence for properties~(a)--(h), verified
by the totality checker, but does not yet connect to a formal Wasm
operational semantics.

\subsection{Tropical and Epistemic Foundations}

Level~11 rests on the tropical semiring $(\mathbb{N} \cup \{\infty\},
\min, +, \infty, 0)$. The Idris~2 formalisation proves the semiring laws:
commutativity of $\oplus$ (\texttt{tropAddComm}), the identity of
$\infty$ for $\oplus$ (\texttt{tropAddIdentity}), and the identity of $0$
for $\otimes$ (\texttt{tropMulIdentity}). Access paths compose via
$\otimes$; the type checker requires a \texttt{CostBounded} witness
proving the total path cost is within a declared bound.

Level~12 rests on epistemic modal logic, specifically the knowledge
operator $K_i$~(``agent~$i$ knows''). Each field in shared memory carries
a monotonic version counter. A module's knowledge is \emph{fresh} when its
observed version equals the field's current version, and \emph{stale}
otherwise. The Idris~2 formalisation proves three key properties:
(1)~a writer always has fresh knowledge of what it wrote
(\texttt{writerKnowsFresh}), (2)~staleness is decidable without axioms
(\texttt{freshOrStale}), and (3)~synchronisation restores freshness
(\texttt{syncRestoresFresh}). All proofs use Nat ordering directly with
zero \texttt{believe\_me}.

% ==========================================================================
\section{Implementation Architecture}
\label{sec:implementation}
% ==========================================================================

\begin{table}[ht]
\centering
\caption{Implementation layers of typed-wasm.}
\label{tab:architecture}
\begin{tabular}{@{}llp{5cm}@{}}
\toprule
\textbf{Layer} & \textbf{Language} & \textbf{Role} \\
\midrule
ABI definitions & Idris~2 & Dependent types: region schemas, access rules, proof obligations \\
FFI bridge & Zig & C-ABI runtime: region management, typed load/store, Wasm codegen \\
Surface parser & ReScript & Parse \texttt{.twasm} syntax to typed AST \\
Proof engine & Idris~2 & Discharge levels 5--10 obligations via totality checker \\
Code generator & Zig $\to$ Wasm & Emit raw Wasm instructions from verified typed access \\
Ecosystem plugin & Rust (TypedQLiser) & Integration as a TypedQLiser target language \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Zig FFI.}
Zig provides native C ABI compatibility, built-in
\texttt{wasm32-freestanding} target support, cross-compilation without
runtime dependencies, and memory-safe defaults. Region handles encode base
offset, schema ID, generation counter (for lifetime tracking), and
ownership flag in a 64-bit value.

\textbf{ReScript parser.}
The surface syntax parser is written in ReScript, consistent with the
VQL-UT parser in the TypedQL ecosystem. ReScript compiles to
JavaScript/Wasm via Deno.

\textbf{TypedQLiser integration.}
typed-wasm implements the \texttt{QueryLanguagePlugin} trait from
TypedQLiser, making Wasm memory a target alongside SQL, GraphQL, and VQL.
This enables end-to-end verification: a database query is type-checked by
TypedQLiser, and the memory region holding its results is type-checked by
typed-wasm.

% ==========================================================================
\section{Evaluation}
\label{sec:evaluation}
% ==========================================================================

We evaluate typed-wasm along three axes: (1)~the compile-time overhead of
type checking region access, (2)~the runtime cost of our approach, and
(3)~a qualitative comparison with alternative safety mechanisms.

\subsection{Type Checking Overhead}

Table~\ref{tab:benchmarks} measures the wall-clock time for type checking
the four example programs shipped with the prototype.  Each program was
checked 100 times; we report the median.  All measurements were taken on
Fedora~43 (kernel~6.19, AMD Ryzen~7 5700U, 16\,GB RAM) using Zig~0.15.2.

\begin{table}[ht]
\centering
\caption{Type checking time for example \texttt{.twasm} programs.}
\label{tab:benchmarks}
\begin{tabular}{@{}lrrrl@{}}
\toprule
\textbf{Program} & \textbf{Regions} & \textbf{Fields} & \textbf{Modules} & \textbf{Check (median)} \\
\midrule
\texttt{01-single-module}       & 1 &  4 & 1 & $<$1\,ms \\
\texttt{02-multi-module}        & 2 &  8 & 2 & $<$1\,ms \\
\texttt{03-ownership-linearity} & 3 & 12 & 2 & 1.2\,ms \\
\texttt{04-ecs-game}            & 5 & 22 & 3 & 2.4\,ms \\
\bottomrule
\end{tabular}
\end{table}

Multi-module schema agreement (the cross-product compatibility check) is
the dominant cost.  For $n$~regions shared across $m$~modules, the worst
case is $O(n \cdot m^2 \cdot f)$ where $f$ is the average field count per
region.  In practice, shared regions are few ($n \le 10$), modules are few
($m \le 5$), and the compatibility check exits early on the first
disagreement.  We did not observe measurable overhead beyond 5\,ms for any
realistic configuration.

\subsection{Runtime Cost}

typed-wasm produces \textbf{zero runtime overhead} by design.  All type
checking occurs at compile time; the emitted Wasm instructions are
identical to those produced without typed-wasm annotations.  This is
achieved through Idris~2 proof erasure
(Section~\ref{sec:formal}): every dependent type, proof term, and linearity
annotation is erased before code generation.

Concretely, the Zig FFI layer emits raw \texttt{i32.load}, \texttt{f64.store},
etc.\@ with computed offsets.  No wrapper functions, no bounds checks, and no
indirection tables are inserted.  The Wasm binary size is identical with and
without typed-wasm annotations.

\subsection{Comparison with Existing Approaches}

Table~\ref{tab:comparison} compares typed-wasm with the approaches
surveyed in Section~\ref{sec:problem} across the five bug classes
(Table~\ref{tab:bugs}) and four architectural properties.

\begin{table}[ht]
\centering
\caption{Coverage comparison: typed-wasm vs.\@ existing approaches.
\checkmark\ = addressed; \textbf{--} = not addressed; $\sim$ = partially addressed.}
\label{tab:comparison}
\begin{tabular}{@{}lccccccccc@{}}
\toprule
& \rotatebox{70}{B1: Type reinterp.} & \rotatebox{70}{B2: Layout disagree.}
& \rotatebox{70}{B3: Null diverge.} & \rotatebox{70}{B4: Lifetime mismatch}
& \rotatebox{70}{B5: Ownership confus.} & \rotatebox{70}{Cross-module}
& \rotatebox{70}{Zero overhead} & \rotatebox{70}{Formal proofs}
& \rotatebox{70}{Progressive} \\
\midrule
Rust borrow ck.       & \checkmark & \textbf{--} & \checkmark & \checkmark & \checkmark & \textbf{--} & \checkmark & $\sim$ & \textbf{--} \\
Wasm validator        & \textbf{--} & \textbf{--} & \textbf{--} & \textbf{--} & \textbf{--} & \textbf{--} & \checkmark & \checkmark & \textbf{--} \\
Wasm GC               & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & $\sim$ & \textbf{--} & $\sim$ & \textbf{--} \\
TAL                   & \checkmark & $\sim$ & \textbf{--} & \textbf{--} & \textbf{--} & \textbf{--} & \checkmark & \checkmark & \textbf{--} \\
MSWasm                & $\sim$ & $\sim$ & \textbf{--} & \textbf{--} & \textbf{--} & $\sim$ & \textbf{--} & $\sim$ & \textbf{--} \\
Component Model       & \checkmark & \checkmark & \checkmark & $\sim$ & $\sim$ & \checkmark & \textbf{--} & \textbf{--} & \textbf{--} \\
\textbf{typed-wasm}   & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

Key observations:

\begin{itemize}
  \item Only typed-wasm covers all five bug classes \emph{and} cross-module
    boundaries simultaneously.  The Rust borrow checker is closest but
    operates within a single language.

  \item The Wasm Component Model is the only other approach with full
    cross-module coverage, but it achieves this by \emph{copying} data across
    boundaries rather than typing shared access --- introducing runtime
    overhead proportional to data size.

  \item typed-wasm is the only approach that is simultaneously
    \emph{progressive} (adoptable at any level) and \emph{formally proven}
    (with dependent types and proof erasure).

  \item MSWasm and typed-wasm are complementary: MSWasm provides spatial
    safety (segment bounds), typed-wasm provides semantic safety (field
    types and schema agreement) within those segments.
\end{itemize}

\subsection{AffineScript TEA: Navigation History as a Linear Type}
\label{sec:affinescript-tea}

% --------------------------------------------------------------------------
% Concrete L10 demonstration: IDApTIK Cadre Router
% --------------------------------------------------------------------------

To validate Level~10 (linearity) in a realistic production context, we
dogfooded typed-wasm's ownership annotations inside
\textsc{IDApTIK}~\cite{idaptik2026}, an asymmetric co-operative
stealth-hacking game built with ReScript and PixiJS.  The experiment
required one concrete claim: \emph{navigation history can be encoded as a
linear type inside a Wasm module}.

\paragraph{The AffineScript TEA architecture.}
AffineScript is a compiler from an affine $\lambda$-calculus front-end to
Wasm~1.0.  It follows The Elm Architecture (\textsc{tea})~\cite{elm2016}:

\[
  \mathtt{update} : \mathtt{Msg} \multimap \mathtt{Model} \multimap \mathtt{Model}
\]

where $\multimap$ is the linear function arrow from QTT
(Section~\ref{sec:formal}).  Every \texttt{update} call consumes the old
model and the incoming message exactly once, producing a fresh model.  The
\texttt{affinescript.ownership} custom section records the linearity kind
(0 = Unrestricted, 1 = Linear) for each function parameter, enabling the
typed-wasm verifier to enforce the invariant statically.

\paragraph{The Cadre Router.}
The \textsc{IDApTIK} front-end requires a bounded back-stack of
at most eight screens.  We compiled a \emph{Cadre Router} Wasm module
(\texttt{router.wasm}, 915~bytes) using the \texttt{affinescript
router-bridge} subcommand.  The router stores a \texttt{RouterModel}
record in linear memory at offset~64:

\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Field}   & \textbf{Offset} & \textbf{Type} & \textbf{Default} \\
\midrule
\texttt{screen\_w}   & +0  & \texttt{i32} & 1280 \\
\texttt{screen\_h}   & +4  & \texttt{i32} & 720 \\
\texttt{stack\_len}  & +8  & \texttt{i32} & 0 \\
\texttt{popup\_tag}  & +12 & \texttt{i32} & $-1$ \\
\texttt{stack\_data} & +16 & \texttt{i32}[8] & $[0 \times 8]$ \\
\bottomrule
\end{tabular}
\end{center}

Five router messages are exported: \texttt{Push}, \texttt{Pop},
\texttt{PresentPopup}, \texttt{DismissPopup}, and \texttt{Resize}.
The \texttt{screen\_tag} parameter of \texttt{Push}, the \texttt{popup\_tag}
of \texttt{PresentPopup}, and both dimension arguments of \texttt{Resize}
are annotated \textbf{Linear} in the \texttt{affinescript.ownership}
custom section.  This means each message is consumed exactly once per
navigation event.  The typed-wasm verifier rejects any module that attempts
to re-send the same tag value without an intervening update.

\paragraph{Integration with ReScript navigation.}
\texttt{Navigation.t} holds an optional \texttt{AffineTEARouterBridge}
(an opaque JS handle to the instantiated Wasm module).  Every
\texttt{showScreenWithTag} and \texttt{presentPopupWithTag} call dispatches
the corresponding \texttt{Push}/\texttt{PresentPopup} message to the router
before the PixiJS transition begins.  The \texttt{goBack} function is a
read-then-reconstruct operation: it calls the linear \texttt{Pop} export,
then reads \texttt{getStackTop()} to identify the previous screen tag, looks
it up in the \texttt{screenRegistry} dictionary (a \texttt{Dict.t<appScreenConstructor>}
populated at startup), and calls \texttt{showScreen} with the recovered
constructor.  All of this occurs without any ReScript-side back-stack ---
the Wasm module \emph{is} the back-stack, stored as a linear resource.

\paragraph{Verification result.}
The router module passes \texttt{WebAssembly.validate}.  The
\texttt{affinescript.ownership} section is checked by the AffineScript
toolchain at code-generation time; the typed-wasm Level~10 checker
additionally verifies at load time that no function parameter marked
\texttt{kind=1} (Linear) is reachable via two distinct call paths within
a single update cycle.  After the full IDApTIK screen suite (TitleScreen
$\to$ CreditsScreen $\to$ \texttt{goBack} $\to$ TitleScreen) was exercised,
the ownership invariant held throughout.

This demonstrates that Level~10 is not merely a theoretical construct: a
real game's navigation history can be encoded as an affine resource inside
a 915-byte Wasm module, with zero runtime overhead and machine-checkable
ownership annotations.

\subsection{ECHIDNA Property Testing}

We additionally validated proof soundness using the ECHIDNA prover oracle
harness (Section~\ref{sec:implementation}).  Seven property-based tests
fuzz the type checker with randomly generated region schemas, field accesses,
and multi-module configurations.  After $10^5$~iterations, no counterexample
was found to any level-1--6 property.  Levels~7--10 properties were tested
with $10^4$~iterations owing to the more complex generation strategy.

% ==========================================================================
\section{Related Work}
\label{sec:related}
% ==========================================================================

\textbf{Typed Assembly Language.}
Morrisett et al.~\cite{morrisett1999} introduced TAL, assigning types to
registers and stack slots to prove progress and preservation for assembly
programs. typed-wasm extends this approach with structured region schemas,
cross-module import/export, and a 12-level progressive framework. TAL
types individual memory cells; typed-wasm types structured multi-field
regions.

\textbf{MSWasm.}
Disselkoen et al.~\cite{disselkoen2022} add memory segments with bounds
checking to Wasm, enforcing spatial safety. typed-wasm builds on this by
adding field-level types within bounded regions, cross-module schema
agreement, effect tracking, and linearity. The two are complementary:
MSWasm ensures access is within bounds; typed-wasm ensures access is also
type-correct within those bounds.

\textbf{Wasm GC Proposal.}
The GC proposal~\cite{wasmgc2024} introduces managed struct and array
types. These are type-safe by construction but apply only to GC-managed
objects, not to linear memory. Systems languages (C, Rust) that manage
their own memory cannot benefit. typed-wasm and Wasm GC are complementary.

\textbf{Wasm Component Model.}
The Component Model~\cite{wasmcomponent2024} types function call interfaces
with automatic marshalling. It addresses isolation-based composition
(copying data across boundaries). typed-wasm addresses shared-memory
composition (typing in-place access). Both are valid architectural choices;
typed-wasm covers the case where copying is too expensive.

\textbf{Rust Ownership.}
Rust's affine type system~\cite{jung2018} prevents use-after-free,
double-free, and data races within Rust code. typed-wasm's Levels~7, 9,
and 10 are deliberately analogous but operate at the Wasm level, covering
cross-language boundaries that Rust cannot reach.

\textbf{Checked C and Typed LLVM.}
Checked C~\cite{elliott2018} adds bounds-checked pointers to C. Both
improve single-language compilation pipelines. typed-wasm operates at the
Wasm convergence point, where multiple source languages meet.

\textbf{Substructural Type Systems.}
Linear types~\cite{wadler1990}, uniqueness types~\cite{barendsen1996}, and
quantitative type theory~\cite{atkey2018} provide the theoretical
foundation for typed-wasm's Levels~7--10. Our contribution is not new type
theory but rather the application of these ideas to the specific problem
of cross-module Wasm memory safety.

\textbf{Tropical Semirings in Type Theory.}
Tropical semirings (min-plus algebras) are widely used in shortest-path
algorithms and automata theory~\cite{simon1988}. Their application to
\emph{type-level cost tracking} for memory access patterns appears to be
novel. Level~11 uses the tropical semiring to bound access costs at
compile time --- a static analogue of database query cost estimation.

\textbf{Epistemic Logic in Verification.}
Epistemic modal logic has been applied to protocol verification and
distributed systems~\cite{fagin1995}. Level~12 adapts the knowledge
operator $K_i$ to shared mutable memory: a module's knowledge of a field
is parameterised by a version counter, and reads require a freshness
proof. This appears to be the first application of epistemic safety to
shared-memory type systems.

% ==========================================================================
\section{Limitations and Future Work}
\label{sec:limitations}
% ==========================================================================

\textbf{Concurrency.}
typed-wasm v0.1 does not address Wasm shared memory with atomics.
Concurrent access to shared regions requires session types or fractional
permissions, which are explored in the TypeQL-Experimental project but not
yet integrated.

\textbf{GC interaction.}
The boundary between linear memory and GC-managed objects is currently
untyped. A \texttt{gcref<T>} field type is conceivable but requires
runtime cooperation beyond static checking.

\textbf{Dynamic layouts.}
Programs that determine memory layout at runtime (JIT compilers, dynamic
object models) cannot use static region declarations without an adaptation
layer.

\textbf{Adoption barrier.}
All participating modules must carry typed-wasm annotations. Automatic
schema inference from \texttt{\#[repr(C)]} attributes, C headers, and Wasm
debug custom sections could lower this barrier.

\textbf{Proof completeness.}
Levels~5--10 depend on the Idris~2 totality checker. Complex arithmetic
bounds may require manual proof assistance. Levels~11--12 are formalised
with proven semiring laws and epistemic properties, but lack Zig FFI
integration and end-to-end testing. Wiring them into the toolchain is
the v1.1 milestone.

\textbf{Mechanised soundness.}
The soundness conjecture (Section~\ref{sec:formal}) is not yet
mechanically verified against a formal Wasm operational semantics.
Connecting the Idris~2 formalisation to an existing Wasm mechanisation
(e.g., WasmCert~\cite{watt2018}) is a priority.

% ==========================================================================
\section{Conclusion}
\label{sec:conclusion}
% ==========================================================================

Wasm's linear memory is the largest untyped shared mutable state surface
in modern software. Multiple languages compile to Wasm; multiple modules
share linear memory; no existing type system covers the boundary between
them.

typed-wasm addresses this by recognising that Wasm memory access has the
same structure as database queries --- projections against a declared
schema. By applying a 12-level progressive type safety framework to
memory regions, typed-wasm provides:

\begin{enumerate}
  \item \textbf{Region schemas} that declare memory layout with field
    names, types, alignment, and invariants.
  \item \textbf{Multi-module schema agreement} that statically verifies
    layout compatibility across independently compiled modules.
  \item \textbf{12 levels of progressive type safety} from instruction
    validity through linearity, tropical cost-tracking, and epistemic
    safety, with formal proofs in Idris~2 and zero runtime overhead
    through proof erasure.
  \item \textbf{A practical architecture} integrating Idris~2, Zig,
    ReScript, and the TypedQLiser ecosystem.
  \item \textbf{A concrete Level~10 demonstration} in a shipping game
    (\textsc{IDApTIK}): navigation history encoded as a linear affine
    resource inside a 915-byte Wasm module, with machine-checked ownership
    annotations and zero runtime overhead
    (Section~\ref{sec:affinescript-tea}).
\end{enumerate}

The implication extends beyond Wasm. Any system where multiple languages
share untyped mutable state --- GraalVM Truffle interop, JNI, FFI bridges,
shared memory IPC --- faces the same class of bugs. typed-wasm demonstrates
that database query type safety is a transferable framework for these
domains.

% ==========================================================================
\appendix
\section{Level Achievement Status}
\label{app:levels}
% ==========================================================================

Table~\ref{tab:levelstatus} reports the implementation status of each
level as of version~1.0.  Levels marked \emph{[sfap]} (``so far as
possible'') have machine-checked proofs in Idris~2 with zero
\texttt{believe\_me}, \texttt{postulate}, or \texttt{assert\_total}.
Full mechanical verification against a formal Wasm operational semantics
(e.g.\ WasmCert-Isabelle~\cite{watt2018}) remains future work.

\begin{table}[ht]
\centering
\caption{Level achievement status (v1.0).}
\label{tab:levelstatus}
\begin{tabular}{@{}rlllll@{}}
\toprule
\textbf{Lv} & \textbf{Name} & \textbf{Proof} & \textbf{Runtime} & \textbf{Tests} & \textbf{Status} \\
\midrule
1  & Instruction     & Region.idr       & Parser      & $10^5$ & E2E \\
2  & Region-binding  & TypedAccess.idr  & Schema      & $10^5$ & E2E \\
3  & Type-compat.    & TypedAccess.idr  & Load/store  & $10^5$ & E2E \\
4  & Null safety     & Pointer.idr      & Ptr kinds   & $10^5$ & E2E \\
5  & Bounds-proof    & Levels.idr       & Bounds chk  & $10^5$ & E2E \\
6  & Result-type     & TypedAccess.idr  & Type flow   & $10^5$ & E2E \\
\midrule
7  & Aliasing        & Pointer.idr      & Erased      & $10^4$ & Proven [sfap] \\
8  & Effects         & Effects.idr      & Erased      & $10^4$ & Proven [sfap] \\
9  & Lifetime        & Lifetime.idr     & Erased      & $10^4$ & Proven [sfap] \\
10 & Linearity       & Linear.idr       & Erased      & $10^4$ & Proven [sfap] \\
\midrule
11 & Tropical cost   & Tropical.idr     & ---         & ---    & Formalised [sfap] \\
12 & Epistemic       & Epistemic.idr    & ---         & ---    & Formalised [sfap] \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Versioning scheme.}
typed-wasm versions track the highest fully-achieved level tier:
v1.0 = L1--10 (proven erasable stack);
v1.1 = L11--12 (tropical + epistemic integrated into toolchain);
v2.0 = L13+ (future levels).

\textbf{``Proven, erased'' (L7--10).}
These levels are verified by the Idris~2 type checker at compile time,
then erased before code generation via QTT.  The emitted Wasm instructions
are identical to hand-written code --- zero runtime overhead.  This is by
design: the proofs catch bugs at compile time and are not needed at runtime.

\textbf{``Formalised'' (L11--12).}
Levels~11 and 12 have Idris~2 type definitions and proof structures with
zero \texttt{believe\_me}, but no Zig FFI implementation.  They are
machine-checked mathematical definitions awaiting toolchain integration.

\textbf{Proof inventory.}
All 13 Idris~2 ABI files (Region, TypedAccess, Levels, Pointer, Effects,
Lifetime, Linear, MultiModule, Proofs, Tropical, Epistemic, and the layout
files ABI, Types, Stdlib) contain zero instances of \texttt{believe\_me},
\texttt{postulate}, or \texttt{assert\_total}.
A full mapping of paper theorems to source locations, including the
Zero-Overhead Property, the tropical semiring laws, and the epistemic
freshness theorems, is in
\texttt{docs/supplementary/proof-inventory.adoc}.
The \texttt{Proofs} module now provides a top-level \texttt{fullCert12}
certificate type requiring concrete L11 (\texttt{AllPairsCosts}) and
L12 (\texttt{Level12Proof}) witnesses, completing the 12-level type
hierarchy.

% ==========================================================================
% Bibliography
% ==========================================================================
\bibliographystyle{ACM-Reference-Format}
\bibliography{typed-wasm}

\end{document}
